---
title: "The Tidymodels Framework"
resource_files:
- appforthat.jpg
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightLines: yes
      highlightStyle: github
      countIncrementalSlides: false
      ratio: '16:9'

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, digits = 4, scipen=999)
library(tidyverse)
library(tidymodels)
library(flair)
library(glmnet)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#26116c",
  text_bold_color = "#fd5e53",
  title_slide_text_color = "#fff8e7",
  background_color = "#fff8e7",
  header_font_google = google_font("Roboto"),
  text_font_google   = google_font("Roboto Condensed"),
  code_font_google   = google_font("Droid Mono")
)
```

```{css, echo = FALSE}
.red{ color: red; }
.blue{ color: blue; }
.huge {
  font-size: 200%;
}
.large {
  font-size: 150%;
}
.tiny {
  font-size: 50%;
}
```


---
class: center, middle, inverse

# A workflow-based approach to modeling

---

## Workflow


0. **Clean** the data

--

1. **Specify** the model(s)

--

2. Create a **recipe** for pre-processing

--

3. Construct a **workflow** establishing the modeling process.

--

4. **Fit** the model

--

5. **Evaluate** the model fit.

---
class: center, middle, inverse

## Clean the data

---
## Clean the data


* Check data summaries and plots for anomalies.

--

* Are there any outliers that may be due to **data entry error**?

--

* Has **R** properly interpreted the *variable types*?

--

* Is the data in **tidy format**?

--

*Cleaning* (as opposed to *pre-processing*) = correcting "mistakes".

---

## Data: The NYC flights dataset

```{r}
library(nycflights13)

set.seed(384)

flights <- flights %>% sample_n(5000)

summary(flights)
```
---

```{r}
flight_data <- 
  flights %>% 
  mutate(
    # Convert the arrival delay to a factor
    arr_delay = ifelse(arr_delay >= 30, "late", "on_time"),
    arr_delay = factor(arr_delay),
    # We will use the date (not date-time) in the recipe below
    date = as.Date(time_hour)
  ) %>% 
  # Include the weather data
  inner_join(weather, by = c("origin", "time_hour")) %>% 
  # Only retain the specific columns we will use
  select(dep_time, flight, origin, dest, air_time, distance, 
         carrier, date, arr_delay, time_hour) %>% 
  # Exclude missing data
  drop_na() %>% 
  # For creating models, it is better to have qualitative columns
  # encoded as factors (instead of character strings)
  mutate_if(is.character, as.factor)
```


---
class: center, middle, inverse

## Specify the model

---
## Specify the model

In `tidymodels`, you need three pieces of information to **specify** a model:

1. The **type** of model.

2. The **engine** - i.e., which R package implementation you will rely on

3. The **mode**: classification or regression.

--

*Optional* information: various parameters of the model

---
## Specify the model

```{r}
lin_mod_spec <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")


logistic_mod_spec <- logistic_reg() %>%
  set_mode("classification") %>%
  set_engine("glm")
  
```

--

**NOTE:** We have not said anything about the dataset here!

---
class: center, middle, inverse

## Recipe for pre-processing

---

## Recipes

**Pre-processing** includes things like:

* **Transforming** variables to be approximately Normal

* Dealing with **missing data**

* **Omitting** certain variables from the model.

* Converting categorical variables to **dummy variables**.

* Adding **interaction terms** to your model

* *(Advanced)* Applying **dimension reduction** to the data.

--

*These are all model-related choices!!!*

---

## Recipes

All a `recipe` **needs** to have is a *formula* and a *dataset*:

```{r}
flights_rec <- 
  recipe(arr_delay ~ ., data = flight_data) 
```

--

Note 1:  The ` ~ .` means "use all available columns" as predictors.  Handy, but dangerous!

--

Note 2: Here `flight_data` is a kind of "example data structure".  It tells the recipe what we can expect, but nothing is computed yet.

---

## Recipes

Now we can add other pre-processing steps to the recipe:

--

Note that the `flight` and `time_hour` columns uniquely identify each flight, and should not be used:

```{r}
flights_rec <- 
  recipe(arr_delay ~ ., data = flight_data) %>% 
  update_role(flight, time_hour, new_role = "ID") 
```

---

Instead of using the `date` column as it stands, let's just note important dates:

```{r}
flights_rec <- 
  recipe(arr_delay ~ ., data = flight_data) %>% 
  update_role(flight, time_hour, new_role = "ID") %>% 
  step_date(date, features = c("dow", "month")) %>%               
  step_holiday(date, holidays = timeDate::listHolidays("US")) %>% 
  step_rm(date)
```

---

Turn all categorical variables into dummy variables, and drop any predictors that have only one observation:

```{r}
flights_rec <- 
  recipe(arr_delay ~ ., data = flight_data) %>% 
  update_role(flight, time_hour, new_role = "ID") %>% 
  step_date(date, features = c("dow", "month")) %>%               
  step_holiday(date, holidays = timeDate::listHolidays("US")) %>% 
  step_rm(date) %>% 
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors())
```

---
## Final Recipe

```{r}
flights_rec
```

---
class: center, middle, inverse

## Make a workflow

---

## Workflow

```{r}
flights_wflow <- workflow() %>%
  add_model(logistic_mod_spec) %>%
  add_recipe(flights_rec)
```

--

**Nothing has yet been calculated!!!**


---
class: center, middle, inverse

## Fit the model

---

## Fitting directly on data

```{r}
model_fit <- flights_wflow %>%
  fit(flight_data)

summary(model_fit)
```

---

```{r}
model_fit %>% pull_workflow_fit()
```

---

```{r}
wflow_fit <- model_fit %>% pull_workflow_fit() 

wflow_fit$fit %>% summary()
```

---

```{r}
model_fit %>% pull_workflow_fit()  %>% tidy()

```



---
class: center, middle

## Evaluate the model

---

## Test/Training

We would like to know how this model will do on *future data*, not just how it fit on the *currently observed* data.

--

Let's set aside 25% of our data to be a "test set":  It will not be used in the **model fitting**, but it will be used for **model evaluation**.

---

## Test/training


```{r}
# Put 3/4 of the data into the training set 
data_split <- initial_split(flight_data, prop = 3/4)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```

---

## First, fit the model on the training data:

```{r}
model_fit <- flights_wflow %>%
  fit(train_data)
```

---

## Then, evaluate the model on the test data:

```{r}
predict(model_fit, test_data)
```

---

```{r}
predict(model_fit, test_data, type = "prob")
```

---

## Comparing predictions to reality

```{r}
flights_pred <- test_data %>% 
        select(arr_delay, time_hour, flight) %>%
  bind_cols(
    predict(model_fit, test_data),
    predict(model_fit, test_data, type = "prob")
  )

flights_pred
```

---
## Comparing predictions to reality

```{r}
flights_pred %>%
  count(.pred_class,
        arr_delay)
```
---

```{r}
flights_pred %>%
  ggplot(aes(x = .pred_late, fill = arr_delay)) +
  geom_density(alpha = 0.5)
```

---

## Evaluating by metrics

```{r}
flights_pred %>%
  accuracy(estimate = .pred_class,
           truth = arr_delay)
```

---

```{r}
flights_pred %>% 
  roc_curve(truth = arr_delay, .pred_late) %>% 
  autoplot()
```

